\bf Supervised Tensor Decomposition with Interactive Side Information

### 2021-04-02 tensor_regress/ICSA_student/latexresource/abstract.tex
submission

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

@@

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Interactive Side Information}
  % blind version
  \author{Jiaxin Hu, Zhuoyan Xu, Chanwoo Lee, and Miaoyan Wang\\
    Department of Statistics, University of Wisconsin-Madison}
  \maketitle
} \fi

@@

\bigskip
\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational political network data. We identify the key global connectivity pattern and pinpoint the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at~\url{https://CRAN.R-project.org/package=tensorregress}.

@@

\newpage
\spacingset{1.5} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Multi-dimensional arrays, known as tensors, are often collected with side information on multiple modes in modern scientific and engineering studies. A popular example is in neuroimaging~\citep{zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in network analysis~\citep{pmlr-v108-berthet20a,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent  friendships. Side information such as people’s demographic information and friendship types are often available. In both examples, it is of keen scientific interest to identify the variation in the data tensor (e.g., brain connectivities, social community patterns) that is affected by available features. These seemingly different scenarios pose a common yet challenging problem for tensor data modeling.

In addition to the aforementioned challenges, many tensor datasets consist of non-Gaussian measurements. Examples include the political interaction dataset \citep{nickel2011three} which measures action counts between countries under various relations, and the brain connectivity network dataset \citep{zhang2018mapping} which is a collection of binary adjacency matrices. Classical tensor decomposition methods are based on minimizing the Frobenius norm of deviation, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, zhao2012higher, yu2016learning,lock2018supervised} to address the tensor regression problem in various forms, such as scalar-to-tensor regression and tensor-response regression. These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap between theory and practice means a great opportunity to modeling paradigms and better capture the complexity in tensor data.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=16cm]{demo.pdf}
\vspace{-.5cm}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatio-temporal growth model.} \label{fig:intro1}
\vspace{-.3cm}
\end{figure*}

We present a general model and associated method for decomposing a data tensor whose entries are from exponential family with interactive side information. We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. Figure~\ref{fig:intro1}b illustrates our model in the special case of order-3 tensors. A low-rank structure is imposed to the conditional mean of tensor observation, where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

Our proposal blends the modeling power of generalized linear model (GLM) and the exploratory capability of tensor dimension reduction in order to take the best out of both worlds. We leverage GLM to allow heteroscedacity due to the mean-variance relationship in the non-Gaussian data. This flexibility is important in practice. Furthermore, our low-rank model on the (transformed) conditional mean tensor effectively mitigates the curse of high dimensionality. In classical GLM, the sample size and feature dimension are well defined; however, in the tensor data analysis, we observe only one realization of an order-$K$ tensor and up to $K$ interactive feature matrices. Both the number of tensor entries and feature dimension grow exponentially in $K$. Dimension reduction is therefore crucial for prediction and interpretability. We establish the statistical convergence of our estimator, and we quantify the gain in prediction through simulations and case studies.

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor, hong2020generalized} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices in our model.

The second line of work studies the tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple}. Our model shares a common ground with earlier approaches, but we provide more efficient solutions to new settings that have more practical significance. As we show in Section~\ref{sec:connection}, the supervised tensor decomposition has an interesting connection with tensor-on-tensor regression. Previous methods~\citep{lock2018tensor} mainly focus on Gaussian tensors. The Frobenius norm used in the objective function is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\citep{raskutti2019convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2019convex}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\textup{Span}(\mM_k)$, that facilitates the identification of interaction effects in features (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of applications.

The third line of work uses side information for various tensor learning tasks, such as for completion~\citep{narita2012tensor} and for recommendation system~\citep{ioannidis2019coupled,farias2019learning}. These methods also study tensors with side information, but they take regularization approaches to penalize predictions that are distant from side information~\cite{lock2018supervised}. One important difference is that their goal is prediction but not parameter estimation. The effects of features and their interactions are not estimated in these data-driven approaches. In contrast, our goal is interpretable prediction, and we estimate factor matrices $\mM_k$ using a model-based approach. Estimating $\mM_k$ allows us to identify sufficient features and the interactions thereof. We numerically compare these two approaches in Section~\ref{sec:simulation}.

The paper is organized as follows. Section~\ref{sec:pre} introduces tensor preliminaries. Section~\ref{sec:model} presents the main model and three motivating examples for supervised tensor decomposition. 
We describe the quasi-likelihood estimation in Section~\ref{sec:est} and provide numerical experiments in Section~\ref{sec:simulation}. In Section~\ref{sec:data}, we present two data applications, one on diffusion tensor imaging data and the other on multi-relational network data.
%and applications to diffusion tensor imaging data and multi-relational social network data in Section~\ref{sec:data}. 
%In Section~\ref{sec:simulation}, we present numerical experiments and assess the performance in comparison to alternative methods. In Section~\ref{sec:data}, we apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We conclude in Section~\ref{sec:con} with discussions about our findings and avenues of future work.

\section{Preliminaries}\label{sec:pre}
We introduce the basic tensor properties~\citep{kolda2009tensor} used in the paper. We use lower-case letters (e.g.,\ $a,b,c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA, \mB, \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC$) for tensors of order three or greater. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY \times_1\mX_1\times \cdots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.$
The tensor Frobenius norm is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$, and the maximum norm is defined $\mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.$
When $a$ is a vector, we use $\vnormSize{}{a}=\langle a,a \rangle^{1/2}$ to denote the vector 2-norm. A higher-order tensor can be reshaped into a lower-order object. We use $\textup{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ to denote the operation that reshapes the tensor along mode $k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\textup{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use $\sigma_{\min}(\cdot)$ and $\sigma_{\max}(\cdot)$ denote the smallest and largest matrix singular values. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+, -, \geq $) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner. 
%We use $f'(\cdot)$, $f''(\cdot)$ to denote the first-order and second-order derivative of a function $f(\cdot)$.

\section{Motivation and model}\label{sec:model}
\subsection{General framework for supervised tensor decomposition}
We begin with a general framework for supervised tensor decomposition and then discuss its implication in three concrete examples. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We propose a multilinear conditional mean model between the data tensor and feature matrices. Assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\quad \text{with}\quad \Theta =\tB\times\{\mX_1,\ldots, \mX_K\} ,
\end{align}
where $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the unknown parameter tensor, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$, and $\times$ denotes the tensor-by-matrix product. The choice of link function is based on the assumed distribution family of tensor entries. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution. In general, dispersion parameters can also be included in the model. Because our main focus is the tensor decomposition under the mean model, we omit the dispersion parameter in this section for ease of presentation.

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$; the adaptation to unknown rank will be addressed in Section~\ref{sec:tuning}. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes. Our goal is to find $\mM_k$ and the corresponding $\tC$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

\subsection{Three examples}
We give three seemingly different examples that can all be formulated as our supervised tensor decomposition model~\eqref{eq:decomp}.
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{gabriel1998generalised} was originally proposed as an example of bilinear model for matrix data, and we adopt its higher-order extension here. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices on each mode. The factors $\mX_k\mM_k$ are sufficient features in the mean model~\eqref{eq:time}. The spatial-temporal model is a special case of our supervised tensor decomposition model~\eqref{eq:decomp}, with features available on each of the three modes.
\end{example}

\begin{example}[Network population model]\label{example:brain}
Network response model~\citep{rabusseau2016low} is recently developed for neuroimaging analysis. The goal is to study the relationship between brain network connectivity pattern and features of individuals. Suppose we have a sample of $n$ observations, $\{(\mY_i, \mx_i)\colon i=1,\ldots,n\}$, where for each individual $i\in[n]$, $\mY_i\in\{0,1\}^{d\times d}$ is the undirected adjacency matrix whose entries indicate presences/absences of connectivities between $d$ brain nodes, and $\mx_i\in\mathbb{R}^p$ is the individual's feature such as age, gender, cognition score, etc. The network-response model  has the conditional mean
\begin{equation}\label{eq:network}
\textup{logit}(\mathbb{E}(\mY_i|\mx_i))=\tB\times_3\mx_i, \quad \text{for }i=1,\ldots,n,
\end{equation}
where $\tB\in \mathbb{R}^{d\times d\times p}$ is a rank-$(r_1,r_1,r_2)$ coefficient tensor, and $\tB$ is assumed to be symmetric in the first two modes.

The model~\eqref{eq:network} is a special case of our supervised tensor decomposition, with feature matrix on the last mode of the tensor. Specifically, we stack the network observations $\{\mY_i\}$ together and obtain an order-3 response tensor $\tY\in\{0,1\}^{d\times d\times n}$. Define a feature matrix $\mX=[\mx_1,\ldots,\mx_n]^T\in\mathbb{R}^{n\times p}$. Then, the model~\eqref{eq:network} has the equivalent representation of supervised tensor decomposition,
\[
\textup{logit}(\mathbb{E}(\tY|\mX))=\tC\times\{\mM,\ \mM,\ \mX\mM'\},
\]
where $\tC\in\mathbb{R}^{r_1\times r_1\times r_2}$ is the core tensor, $\mM\in\mathbb{R}^{d\times r}$ is the dimension reduction matrix on the first two modes, and $\mM'\in\mathbb{R}^{p\times r_2}$ is for the last mode.  \end{example}

 \begin{example}[Dyadic data with node attributes] Dyadic dataset consists of measurements on pairs of objects. Common examples include graphs and networks. Let $\tG=(V,E)$ denote a graph, where $V=[d]$ is the node set of the graph, and $E\subset V\times V$ is the edge set. Suppose that we also observe feature vector $\mx_i\in\mathbb{R}^p$ associated to each node $i\in V$. A probabilistic model on the graph $\tG=(V,E)$ can be described by the following matrix regression. The edge connects the two vertices $i$ and $j$ independently of other pairs, and the probability of connection is modeled as
\begin{equation}\label{eq:edge}
 \textup{logit}\left(\mathbb{P}\left((i,j)\in E\right)\right)=\mx^T_i\mB\mx_j=\langle \mB,\ \mx^T_i\mx_j\rangle,
 \end{equation}
 where $\mB\in\mathbb{R}^{p\times p}$ is a symmetric rank-$r$ matrix. The low-rankness in $\mB$ has demonstrated its success in modeling transitivity, balance, and communities in  networks~\citep{hoff2005bilinear}. We show that our supervised tensor decompostion~\eqref{eq:decomp} also incorporates the graph model as a special case. Let $\tY=\entry{y_{ij}}$ be a binary matrix where $y_{ij}=\mathds{1}_{(i,j)\in E}$. Define $\mX=[\mx_1,\ldots,\mx_n]^T\in\mathbb{R}^{n\times p}$. Then, the graph model~\eqref{eq:edge} can be expressed as
 \[
 \textup{logit}(\mathbb{E}(\mY|\mX))=\mC\times\{\mX\mM,\ \mX\mM\}, 
  \]
  where $\mC\in\mathbb{R}^{r\times r}, \mM\in\mathbb{R}^{p\times r}$ are from the singular value decomposition of $\mB=\mM\mC\mM^T$. 
  \end{example}

%In the above three examples and many other studies, researchers are interested in uncovering the variation in the data tensor that can be explained by features. The coefficient tensor $\tB$ in model~\eqref{eq:tensormodel} serves this goal by collecting the feature effects. To encourage the sharing among effects, decomposition~\eqref{eq:factor} assumes that the feature effects $\tB$ lie in a low-dimensional parameter space. The low-rank assumption is plausible in many scientific applications. In the Example~\ref{example:brain} of brain imaging analysis, for instance, it is often believed that the brain nodes can be partitioned into several regions, and the person's attributes affect the connectivity in a similar way within a region.

In the above three examples and many other studies, researchers are interested in uncovering the variation in the data tensor that can be explained by features. Our supervised tensor decomposition~\eqref{eq:decomp} allows arbitrary numbers of feature matrices. When certain mode $k$ has no side information, we set $\mX_k=\mI_{d_k}$ in the model~\eqref{eq:decomp}. In particular, our model~\eqref{eq:decomp} reduces to classical unsupervised tensor decomposition~\citep{de2000multilinear,hong2020generalized} when no side information is available; i.e., $\mX_k=\mI_{d_k}$ for all $k\in[K]$.

\subsection{Connection to sufficient dimension reduction and tensor-on-tensor regression}\label{sec:connection}
One important implication of our method is that we allow high-dimensionality in both tensor dimension $d_k$ and feature dimension $p_k$. Recall that the model rank $r_k$ is typically smaller than $d_k$ and $p_k$. In such a case, the matrices $\mM_k$ serve the role of simultaneous dimension reduction of the data tensor and features. The role of the sufficient features $\mX_k\mM_k$ can be seen in the following two conditional independence assumptions in model~\eqref{eq:decomp},

\begin{align}
\tY \ci \{\mX_k\}&\ \big|\ \{\mX_k\mM_k\} \ \ \text{(independence between the tensor and features)},\\
y_{i_1,\ldots,i_K} \ci y_{i'_1,\ldots,i'_K} & \ \big|\ \{\mX_k\mM_k\}  \ \ \text{(independence within the tensor)},
\end{align}
where the second line holds for all $(i_1,\ldots,i_K)\neq(i'_1,\ldots,i'_K) \in[d_1]\times\cdots\times[d_K]$, and $\ci$ denotes the independence. The first property highlights the ``decorrelation'' role of $\mM_k$, in the same spirit as the sufficient dimension reduction~\citep{adragni2009sufficient} in supervised learning, whereas the second property highlights the tensor dimension reduction in the usual unsupervised sense (consider, for example, $\mX_k=\mathbf{I}_{d_k}$ for all $k\in[K]$).

Our model also has a close connection with tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple}. Specifically, model~\eqref{eq:decomp} can be viewed as a multivariate regression model, where the response is the vectorized tensor and the covariates are interactions between sufficient features across modes. We take an order-3 tensor under the Gaussian model for illustration. Let $\mX,\mZ,\mW$ denote the feature matrix on mode $k=1, 2, 3$, respectively. Suppose that each mode has two-dimensional sufficient features, denoted $\mM_1\mX=[\mx_1,\mx_2]$, $\mM_2\mZ=[\mz_1,\mz_2]$, $\mM_3\mW=[\mw_1,\mw_2]$. Here $\mx_1,\mx_2,\ldots,\mw_1, \mw_2$ are column vectors. Then the model~\eqref{eq:decomp} is a regression model with across-mode interactions,
\begin{equation}\label{eq:interaction}
\mathbb{E}(y_{ijk}|\mX,\mZ,\mW)=c_{111}\mx_{1i}\mz_{1j}\mw_{1k}+c_{121}\mx_{i1}\mz_{j2}\mw_{k1}+\cdots+c_{221}\mx_{2i}\mz_{2j}\mw_{1k}+c_{222}\mx_{2i}\mz_{2j}\mw_{2k},
\end{equation}
where $\entry{c_{ijk}} \in \mathbb{R}^{2\times 2\times 3}$ are unknown interaction effects, $\mx_{1i}$ denotes the $i$-th entry in the feature vector $\mx_1$, and similar notations apply to other features. Note that lower-order interactions are naturally incorporated in~\eqref{eq:interaction} if we include an intercept column in the sufficient feature matrices. Model~\eqref{eq:interaction} shows the benefit of our supervised tensor decomposition for identifying across-mode interactions.

\section{Estimation}\label{sec:est}

\subsection{Rank-constrained M-estimator}
We develop a likelihood-based procedure to estimate $\tC$ and $\mM_k$ in~\eqref{eq:decomp}. We adopt the exponential family as a flexible framework for different data types. In a classical generalized linear model with a scalar response $y$ and feature $\mx$, the density is expressed as
\[
p(y|\mx, \boldsymbol{\beta})=c(y,\phi)\exp\left(\frac{y\theta- b(\theta)}{\phi}\right)\ \text{with}\ \theta=\boldsymbol{\beta}^T\mx,
\]
where $b(\cdot)$ is a known function, $\theta$ is the linear predictor, $\phi>0$ is the dispersion parameter, and $c(\cdot)$ is a known normalizing function. The choice of link functions depends on the data types and on the observation domain of $y$, denoted $\mathbb{Y}$. For example, the observation domain is $\mathbb{Y}=\mathbb{R}$ for continuous data, $\mathbb{Y}=\mathbb{N}$ for count data, and  $\mathbb{Y}=\{0,1\}$ for binary data. The canonical link function $f$ is chosen to be $f(\cdot)=b'(\cdot)$, the first-order derivative of $b(\cdot)$. Table~1 summarizes the canonical link functions for common types of distributions.

\begin{table}[htb]
\centering
\begin{tabular}{c|ccc}
Data type &Gaussian & Poisson& Bernoulli\\
\hline
Domain $\mathbb{Y}$& $\mathbb{R}$&$\mathbb{N}$&$\{0,1\}$\\
 $b(\theta)$&$\theta^2/2$& $\exp(\theta)$&$\log (1+\exp(\theta))$\\
 link $f(\theta)$&$\theta$&$\exp(\theta)$&$(1+\exp(-\theta))^{-1}$
\end{tabular}
\vspace{-.2cm}
\caption{Canonical links for common distributions.}\label{table:link}
\end{table}

In our context, we model the tensor entries $y_{i_1,\ldots,i_K}$, conditional on $\theta_{i_1,\ldots,i_K}$, as independent draws from an exponential family. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\}.
\end{equation}
We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where the parameter space $\tP$ is
\[
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
\]
The maximum norm constraint on the linear predictor $\Theta$ is a technical condition to avoid the divergence in the non-Gaussian variance.

\subsection{Alternating optimization} \label{subsec:alg}
We propose an alternating optimization algorithm to solve~\eqref{eq:MLE}. 
The decision variables in the objective function~\eqref{eq:MLE} consist of $K+1$ blocks of variables, one for the core tensor $\tC$ and $K$ for the factor matrices $\mM_k$. We notice that, if any $K$ out of the $K+1$ blocks of variables are known, then the optimization reduces to a simple GLM with respect to the last block of variables. This observation leads to an iterative updating scheme for one block at a time while keeping others fixed. After each iteration, we rescale the core tensor $\tC^{(t+1)}$ subject to the maximum norm constraint. The full algorithm is described in Algorithm~\ref{alg:B}.

\begin{algorithm}[!h]
\caption{Supervised Tensor Decomposition with Interactive Side Information}\label{alg:B}
\begin{algorithmic}[1]
\INPUT Response tensor $\tY\in \mathbb{R}^{d_1\times \cdots \times d_K}$, feature matrices $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,\ldots,K$, target Tucker rank $\mr=(r_1,\ldots,r_K)$, link function $f$, maximum norm bound $\alpha$
\OUTPUT Estimated core tensor $\hat \tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ and factor matrices $\hat \mM_k\in\mathbb{R}^{p_k\times r_k}$. 
\State Random initialization of the core tensor $\tC$ and factor matrices $\mM_k$. 
\While{Do until convergence}
\For { $k=1$ to $K$}
\State Obtain the factor matrix $\tilde \mM_k\in\mathbb{R}^{p_k\times r_k}$ by a GLM with link function $f$.
\State Perform QR factorization $\tilde \mM_k=\mQ\mR$.
\State Update $\mM_k\leftarrow \mQ$ and core tensor $\tC\leftarrow \tC\times_k \mR$.
\EndFor
\State Update the core tensor $\tC$ by solving a GLM with $\textup{vec}(\tY)$ as response, $\otimes_{k=1}^K[ \mX_k\mM_k]$ as features, and $f$ as link function. Here $\otimes$ denotes the Kronecker product of matrices. 
\State Rescale the core tensor $\tC$ subject to the maximum norm constraint. 
\EndWhile
\end{algorithmic}
\end{algorithm}

The optimization~\eqref{eq:MLE} is a non-convex problem due to the non-convexity in the feasible set $\tP$. Under mild conditions, our algorithm enjoys global convergence; i.e.\ any sequence of iterates generated by the alternating algorithm converges to a stationary point of $\tL_\tY(\cdot)$ module orthogonal transformation (see Supplementary Notes, Section~\ref{sec:SAlgorithm}). In theory, global optimality of non-convex optimization is often challenging to obtain. Fortunately, we will show in Section~\ref{subsec:statprob} that the desired statistical performance holds for all local optimizers satisfying $\tL_{\tY}(\hat \tC, \hat \mM_1,\ldots,\hat \mM_K)\geq \tL_{\tY} (\tC_{\textup{true}},\mM_{1,\textup{true}},\ldots,\mM_{K,\textup{true}})$, where the subscript ``true'' denotes the true parameter in model~\eqref{eq:decomp}. This result indicates the global optimality is not necessarily a serious concern in our context, as long as the convergent objective is large enough. In the experiments we considered, we find that Algorithm~\ref{alg:B} typically gives satisfactory convergent points upon random initialization. Figure~\ref{fig:loglike} shows the trajectories of objective function for order-3 tensors based on model~\eqref{eq:decomp}, where $d_k = d \in\{25,30\}$, $p_k = 0.4d$, $r_k = r\in\{3,6\}$ for $ k = 1,2,3$. We consider input tensors with Gaussian, Bernoulli, and Poisson entries. Under all combinations of the dimension $d$, rank $r$, and type of the entries, Algorithm~\ref{alg:B} converges quickly in a few iterations, and the objective values at convergent points are close to or larger than the value at true parameters.  The time complexity is provided in Supplementary Notes.

\begin{figure}[t]
\centering
\includegraphics[width=15cm]{loglike.pdf}
\vspace{-.25cm}
\caption{Trajectory of the objective function with various dimension $d$ and rank $r$ under (a) Gaussian (b) Bernoulli (c) Poisson models. The dashed line represents the objective value at true parameters. }\label{fig:loglike}
\vspace{-.3cm}
\end{figure}

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1in}%
\addtolength{\topmargin}{-.8in}%

@@

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Interactive Side Information}
  % blind version
 \author{
    Jiaxin Hu\\
   \texttt{jhu267@wisc.edu}\\
   \texttt{Tel:6089600989}
  \and
    Chanwoo Lee\\
   \texttt{chanwoo.lee@wisc.edu}\\
   \texttt{Tel:6085569906}
    \and
    Miaoyan Wang\\
    \texttt{miaoyan.wang@wisc.edu}\\
    \texttt{Tel:6082653990}\\
    \\
    Department of Statistics, University of Wisconsin-Madison\\
    \texttt{1300 University Ave,Madison,WI 53706,USA}

}

  %\author{Jiaxin Hu, Chanwoo Lee, and Miaoyan Wang\\
   % Department of Statistics, University of Wisconsin-Madison}
   \date{}
  \maketitle
} \fi

@@

\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational political network data. We identify the key global connectivity pattern and pinpoint the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at~\url{https://CRAN.R-project.org/package=tensorregress}.

@@
